Co-citations in context

### 2019-05-22 cocit_r.tex
more tweaks and Chacko subsection

\section{Introduction}
Citation and network analysis of  scientific literature reveals information on semantic relationships between publications, collaboration between scientists, and the practice of citation itself~\citep{garfield_citation_1955,de_solla_price_networks_1965,newman_structure_2001,Shi:2010:CHI:1816123.1816131,patience_pmid28560354}. Co-citation, the frequency with which two documents are cited together in other documents provides additional insights, including the identification of semantically related documents, fields, idea, and specialization in science~\citep{small_co-citation_1973,marshakova-shaikevich_co-citation_1973}.

\cite{uzzi_atypical_2013} used a novel approach for co-citation analysis of 17.9 million articles and their cited references from the Web of Science (WoS)  to characterize a subset of highly cited articles with respect to both novel and conventional combinations of prior research. Observed co-citation frequencies, mapped to journals, were computed across the dataset and normalized (shifted and scaled) to expected values generated by Monte Carlo simulations under a random graph model. These normalized journal pair frequencies were termed \emph{z-scores} (Materials and Methods). Thus, every article was associated with multiple z-scores. For each article, positional statistics of these z-scores were calculated to describe conventionality: high conventionality (HC) if the median z-score for an article was greater than the median of median z-scores of all articles and low conventionality for the inverse (LC). Similarly,  an article was deemed to have high novelty (HN) if the tenth percentile of its z-scores  was negative and low novelty (LN) for the inverse. Accordingly, each article was labeled with respect to conventionality and novelty, e.g, HCHN (denoting that the article exhibits high conventionality and high novelty), with all four combinations being possible. 
\cite{uzzi_atypical_2013} observed that HCHN articles were twice as likely to be found in highly cited papers, suggesting that novel combinations of ideas flavoring a body of conventional thought were a recipe for impact across the scientific literature.

Key to Uzzi {\em et al.}, however, is their random graph model and its underlying assumptions. A careful examination of the model (as defined by the Monte Carlo simulation) reveals that all substitutions of references are equiprobable, which means that a reference in an article can be substituted by any other reference randomly chosen from those published in the same year, irrespective of subject matter and citation count. For example, the model permits substitution of a reference in quantum physics with equal probability by a reference in classical literature, evolutionary biology, or anthropology. Such substitutions poorly model documented citation behavior~\citep{wallace_lariviere_gingras_2012,moed_measuring_2010,klavans_research_2017,garfield_1979}. In addition, under this random model,  a reference cited over 100 times in a given year is selected with the same probability as a reference cited only once, which is also inconsistent with the citation frequency distribution (Stringer, Perline).  Accordingly, the expected value calculations generated by the \cite{uzzi_atypical_2013} simulations and used in characterizing journal pairs in terms of conventionality and novelty  can be reasonably questioned on grounds of model mis-specification.

A follow-up study by \cite{boyack_vs_uzzi_2014}  explored the impact of discipline and journal effects on the definition of conventionality and novelty.  While their study had some methodological differences from \cite{uzzi_atypical_2013} in the use of Scopus data and a $\chi^2$ calculation for expected values, Boyack and Klavans find the same basic trend that HCHN is more probably in highly cited papers. However, they note that ``only 64.4\%  of  243  WoS  subject  categories'' in the Uzzi {\em et al.} study met the criterion of having the highest probability of hit papers in the HCHN category.  Further, they observed that journals vary widely in terms of size and influence and that 20 journals accounted for nearly 15\% of co-citations in their measurements. Lastly, they noted that three multidisciplinary journals accounted for 9.4\% of all atypical combinations, suggesting strong effects from both disciplines and journals that were not reported by Uzzi et al.

Despite different methods used to generate expected values, both \cite{uzzi_atypical_2013} and \cite{boyack_vs_uzzi_2014}  measured observed frequencies across the scientific literature and used these pan-literature z-scores to examine disciplinary subsets. In contrast, we chose instead to first construct semantically related sets of documents (disciplinary networks) and measure observed and expected frequencies within these networks.

Accordingly, we used keyword searches of the scientific literature to create three citation networks themed around broad search terms. Within these disciplinary frameworks, we calculated expected values using a random graph model that accounted for existing citation frequencies and an efficient simulation algorithm that permitted a substantially greater number of simulations compared to Uzzi.  We hypothesized that our approach would  reduce model misspecification and better simulate citation practice, in turn leading to different and more relevant conclusions. Our study on these semantically-themed citation networks reveals significantly different patterns of conventionality and novelty between  citation networks and disciplines that challenges the conclusion that HCHN articles have a greater probability of being represented in highly cited publications. 
Instead, we conclude that conventional thought as commonly defined in our study, 
in \cite{uzzi_atypical_2013}, and  \cite{boyack_vs_uzzi_2014}, is more likely to drive higher citations..
 \emph{these are just filler conclusions that must be carefully rewritten and added to}.

\section{Materials and Methods}
\emph{Bibliographic data} We have previously developed ERNIE, an open source knowledge platform into which we parse the Web of Science (WoS) Core Collection~\citep{Keserci371955}. WoS data stored in ERNIE spans the period 1900-2019 and consists of over 72 million publications. For this study, we generated an analytical dataset from years 1985 to 2005. The total number of publications in this dataset was just over 25 million publications (25,134,073).  For each of these years, we further restricted analysis to those of type Article. Since WoS data also contains incomplete references or references that point at other indexes,  we also considered only those references for which there were complete records (Table 1). For example, WoS data for year 2005 contained 1,753,174 publications, which after restricting to type Article and considering only those references described above resulted in 916,573 publications, 6,095,594 unique references (set of references), and 17,167,347 total references (multiset of references). Given consistent trends in the data, we analyzed the two boundary years (1985 and 2005) and the mid-point (1995) performing 1,000 simulations for each dataset. \\
\emph{Disciplinary datasets} We constructed three disciplinary datasets based on keyword searches. (i) immunology (ii) metabolism (iii) applied physics. For the first two, rooted in biomedical research, we searched Pubmed for the term `immunology' or `metabolism' in the years 1985, 1995, and 2005. Pubmed IDs (pmids) returned were converted to WoS IDs (wos\_ids). For the applied physics dataset, we directly searched article level labels in WoS for `applied physics'. \\
\emph{Normalization of observed and expected values} Building upon the work of Uzzi {\em et al.}, all ${n \choose 2}$ reference pairs were generated for each publication, where $n$ is the number of cited references in the publication. These reference pairs were then mapped to the journals they were published in using ISSN numbers,  creating journal pairs. Where multiple ISSN numbers exist for a journal, the most frequently used one in the WoS was assigned to the journal. In addition, publications containing less than two references were discarded. Journal pair frequencies were summed up across the dataset to create observed frequencies $(F_{obs})$. In contrast to the preceding study~\citep{uzzi_atypical_2013}, we generated 1,000 rather than 10 null models for each dataset by randomly shuffling references while preserving the number of publications, the number of references in each publication, and the frequency with which these references were cited within the year of interest. Expected values $(F_{exp})$ were generated by averaging the result of 1,000 simulations. z-scores were calculated for each journal-pair using the formula $(F_{obs} - F_{exp})/\sigma$ where $\sigma$
 is the standard deviation of the frequencies generated by simulation. As a result of these calculations, each publication becomes associated with a set of z-scores corresponding to the journal pairs derived from pairwise combinations of its cited references and positional statistics (quantiles) of z-scores were calculated for each publication. Publications were also labeled according to conventionality and novelty. (i) HC if the median z-score exceeded the median of median z-scores for all publications. LC if the median z-score was equal to or less than the median of median z-scores for all publications (ii) HN if the tenth percentile of z-scores for a publication was less than zero. LN if the tenth percentile of z-scores for a publication was greater than zero.

@@

\section{Results}

(i) disciplinary networks are different
(ii) background makes a difference
(iii) 
(iv)

@@

\acknowledgments
We deeply thank Kevin Boyack and Dick Klavans for critically constructive discussions. We thank the authors of Uzzi et al. (2013) for sharing their Python simulation code. Research and development reported in this publication was partially supported by Federal funds from the National Institute on Drug Abuse, National Institutes of Health, US Department of Health and Human Services, under Contract Nos. HHSN271201700053C (N43DA-17-1216) and HHSN271201800040C (N44DA-18-1216). The content of this publication is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health. TW receives funding from the Grainger Foundation. All the code used in this study is freely available from a Github site. Access to bibliographic data requires a license from Clarivate Analytics, which had no role in 
funding, experimental design, review of results, and conclusions presented.

>>>>

\section{Introduction}
Citation and network analysis of  scientific literature reveals information on semantic relationships between publications, collaboration between scientists, and the practice of citation itself~\citep{garfield_citation_1955,de_solla_price_networks_1965,newman_structure_2001,Shi:2010:CHI:1816123.1816131,patience_pmid28560354}. Co-citation, the frequency with which two documents are cited together in other documents provides additional insights, including the identification of semantically related documents, fields, idea, and specialization in science~\citep{small_co-citation_1973,marshakova-shaikevich_co-citation_1973,boyack_co-citation_2010}.

\citep{uzzi_atypical_2013} used a novel approach for co-citation analysis of 17.9 million articles and their cited references from the Web of Science (WoS)  to characterize a subset of highly cited articles with respect to both novel and conventional combinations of prior research. Observed co-citation frequencies, mapped to journals, were computed across the dataset and normalized (shifted and scaled) to expected values generated by Monte Carlo simulations under a random graph model. These normalized journal pair frequencies were termed \emph{z-scores} (Materials and Methods). Thus, every article was associated with multiple z-scores. For each article, positional statistics of these z-scores were calculated to describe conventionality: high conventionality (HC) if the median z-score for an article was greater than the median of median z-scores of all articles and low conventionality for the inverse (LC). Similarly,  an article was deemed to have high novelty (HN) if the tenth percentile of its z-scores  was negative and low novelty (LN) for the inverse. Accordingly, each article was labeled with respect to conventionality and novelty, e.g, HCHN (denoting that the article exhibits high conventionality and high novelty), with all four combinations being possible. 
\citep{uzzi_atypical_2013} observed that HCHN articles were twice as likely to be found in highly cited papers, suggesting that novel combinations of ideas flavoring a body of conventional thought were a recipe for impact across the scientific literature.

Key to Uzzi {\em et al.}, however, is their random graph model and its underlying assumptions. A careful examination of the model (as defined by the Monte Carlo simulation) reveals that all substitutions of references are equiprobable, which means that a reference in an article can be substituted by any other reference randomly chosen from those published in the same year, irrespective of subject matter and citation count. For example, the model permits substitution of a reference in quantum physics with equal probability by a reference in classical literature, evolutionary biology, or anthropology. Such substitutions poorly model documented citation behavior~\citep{wallace_lariviere_gingras_2012,moed_measuring_2010,klavans_research_2017,garfield_1979}. In addition, under this random model,  a reference cited over 100 times in a given year is selected with the same probability as a reference cited only once, which is inconsistent with the frequency distribution of citations characterized as power law or lognormal~\citep{stringer_statistical_2010,perline_strong_2005}.  Accordingly, the expected value calculations generated by the \citep{uzzi_atypical_2013} simulations and used in characterizing journal pairs in terms of conventionality and novelty  can be reasonably questioned on grounds of model mis-specification.

A follow-up study by \citep{boyack_vs_uzzi_2014}  explored the impact of discipline and journal effects on the definition of conventionality and novelty.  While their study had some methodological differences from \citep{uzzi_atypical_2013} in the use of Scopus data and a $\chi^2$ calculation for expected values, Boyack and Klavans find the same basic trend that HCHN is more probably in highly cited papers. However, they note that ``only 64.4\%  of  243  WoS  subject  categories'' in the Uzzi {\em et al.} study met the criterion of having the highest probability of hit papers in the HCHN category.  Further, they observed that journals vary widely in terms of size and influence and that 20 journals accounted for nearly 15\% of co-citations in their measurements. Lastly, they noted that three multidisciplinary journals accounted for 9.4\% of all atypical combinations, suggesting strong effects from both disciplines and journals that were not reported by Uzzi et al.

Despite different methods used to generate expected values, both \citep{uzzi_atypical_2013} and \citep{boyack_vs_uzzi_2014}  measured observed frequencies across the scientific literature without disciplinary constraints and subsequently used normalized frequencies to examine disciplinary subsets. In contrast, we chose instead to first construct semantically related sets of documents (disciplinary networks) to and measure observed and expected frequencies within these networks.

Accordingly, we used keyword searches of the scientific literature to create three citation networks themed around broad search terms. Within these disciplinary frameworks, we calculated expected values using a random graph model that accounted for existing citation frequencies and an efficient Monte Carlo simulation algorithm that permitted a substantially greater number of simulations. We hypothesized that our approach would  reduce model misspecification and better simulate citation practice, in turn leading to different and more relevant conclusions. Our study on these semantically-themed citation networks reveals significantly different patterns of conventionality and novelty between  citation networks and disciplines that challenges the conclusion that HCHN articles have a greater probability of being represented in highly cited publications. 
Instead, we conclude that conventional thought as commonly defined in our study, 
in \citep{uzzi_atypical_2013}, and  \citep{boyack_vs_uzzi_2014}, is more likely to drive higher citations..
 \emph{these are just filler conclusions that must be carefully rewritten and added to- need Jim's conclusions to be stated here.}.

\section{Materials and Methods}
\emph{Bibliographic data} We have previously developed ERNIE, an open source knowledge platform into which we parse the Web of Science (WoS) Core Collection~\citep{Keserci371955}. WoS data stored in ERNIE spans the period 1900-2019 and consists of over 72 million publications. For this study, we generated an analytical dataset from years 1985 to 2005. The total number of publications in this dataset was just over 25 million publications (25,134,073).  For each of these years, we further restricted analysis to those of type Article. Since WoS data also contains incomplete references or references that point at other indexes,  we also considered only those references for which there were complete records (Table 1). For example, WoS data for year 2005 contained 1,753,174 publications, which after restricting to type Article and considering only those references described above resulted in 916,573 publications, 6,095,594 unique references (set of references), and 17,167,347 total references (multiset of references). Given consistent trends in the data, we analyzed the two boundary years (1985 and 2005) and the mid-point (1995) performing 1,000 simulations for each dataset. \\
\emph{Disciplinary datasets} We constructed three disciplinary datasets based on keyword searches. (i) immunology (ii) metabolism (iii) applied physics. For the first two, rooted in biomedical research, we searched Pubmed for the term `immunology' or `metabolism' in the years 1985, 1995, and 2005. Pubmed IDs (pmids) returned were matched to WoS IDs (wos\_ids) and used to retrieve relevant articles. For the applied physics dataset, we directly searched article level labels in WoS for `applied physics'. \\
\emph{Normalization of observed and expected values} Building upon prior work~\cite{uzzi_atypical_2013}, all ${n \choose 2}$ reference pairs were generated for each publication, where $n$ is the number of cited references in the publication. These reference pairs were then mapped to the journals they were published in using ISSN numbers,  creating journal pairs. Where multiple ISSN numbers exist for a journal, the most frequently used one in the WoS was assigned to the journal. In addition, publications containing less than two references were discarded. Journal pair frequencies were summed up across the dataset to create observed frequencies $(F_{obs})$. In contrast to the preceding study~\citep{uzzi_atypical_2013}, we generated 1,000 rather than 10 null models for each dataset by randomly shuffling references while preserving the number of publications, the number of references in each publication, and the frequency with which these references were cited within the year of interest. Expected values $(F_{exp})$ were generated by averaging the result of 1,000 simulations. z-scores were calculated for each journal-pair using the formula $(F_{obs} - F_{exp})/\sigma$ where $\sigma$
 is the standard deviation of the frequencies generated by simulation. As a result of these calculations, each publication becomes associated with a set of z-scores corresponding to the journal pairs derived from pairwise combinations of its cited references and positional statistics (quantiles) of z-scores were calculated for each publication. Publications were also labeled according to conventionality and novelty. (i) HC if the median z-score exceeded the median of median z-scores for all publications. LC if the median z-score was equal to or less than the median of median z-scores for all publications (ii) HN if the tenth percentile of z-scores for a publication was less than zero. LN if the tenth percentile of z-scores for a publication was greater than zero.

@@

\section{Results}
\subsection{Chacko}
(i) disciplinary networks are different
(ii) background makes a difference
(iii) 
(iv)

@@

\acknowledgments
We are grateful to Kevin Boyack and Dick Klavans for critically constructive discussions. We thank the authors of Uzzi et al. (2013) for sharing their Python simulation code. Research and development reported in this publication was partially supported by Federal funds from the National Institute on Drug Abuse, National Institutes of Health, US Department of Health and Human Services, under Contract Nos. HHSN271201700053C (N43DA-17-1216) and HHSN271201800040C (N44DA-18-1216). The content of this publication is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health. TW receives funding from the Grainger Foundation. All the code used in this study is freely available from a Github site. Access to the bibliographic data analyzed in this study requires a license from Clarivate Analytics, which had no role in funding, experimental design, review of results, and conclusions presented.
